---
title: "Working With Strings, a Few Tricks"
subtitle: "*Basics of R for Data Science*"
include-in-header: ../assets/header.html
format: 
  html:
    self-contained: true
    toc: true   
    toc-depth: 4  
    toc-expand: 4
    code-fold: false 
    message: false
    css: "../assets/custom.css"
editor: visual
---

#### Uppercase and lowercase with `toupper()` and `tolower()`

```{r}
text = "This is a String"

toupper(text)

tolower(text)
```

#### Replace a character with another character in a string using `gsub()`
```{r}
text = "this is a string"

gsub("t", "X", text)
```

#### Removing all punctuation 
```{r}
text = "This is a string. For some reason, there is some punctuation."

gsub("[[:punct:]]", "", text)
```

#### Removing all digits 
```{r}
text = "in this string231 there are some 1241useless digits 443perhaps because 3434someone copy and pasted 985from a pdf"

gsub("[[:digit:]]", "", text)
```

#### Splitting a string into a vector of words 
```{r}
text = "this is a string"

unlist(strsplit(text, split=" "))
```

*Note:* `strplit()` actually does the job, but it returns a list; if you want a vector for convenience, use `unlist()` on the output list

#### Removing a few cases from a vector
```{r}
myVect = c("this", "is", "a", "vector")

myVect[!myVect %in% c("is", "a")]
```

#### Counting characters in strings
```{r}
myVect = c("this", "is", "a", "vector")

nchar(myVect)
```

#### Turning a vector of strings into a single string

```{r}
myVect = c("this", "is", "a", "vector")

paste(myVect, collapse=" ")
```



